Sparse spectral and $p$-finite element methods for partial differential equations on disk slices and trapeziums

### 2019-05-29 papers/halfdisk/halfdisk.tex
draft

\newcommand{\hdop}{H}
\newcommand{\bighdop}{\mathbb{\hdop}}
\newcommand{\hdopnk}{\hdop_{n,k}}
\newcommand{\hdopab}{\hdop^{(a,b)}}
\newcommand{\hdopnkab}{\hdop_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\hdopmj}{\hdop_{m,j}}
\newcommand{\hdopmjab}{\hdop_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bighdopab}{\bighdop^{(a,b)}}
\newcommand{\Dnt}{D^\top_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\hdopii}{\hdop^{(1,1)}}
\newcommand{\bighdopii}{{\mathbb{\hdop}^{(1,1)}}}
\newcommand{\hdopoo}{\hdop^{(0,0)}}
\newcommand{\bighdopoo}{{\mathbb{\hdop}^{(0,0)}}}
\newcommand{\hdopoi}{\hdop^{(0,1)}}
\newcommand{\hdopio}{\hdop^{(1,0)}}
\newcommand{\jac}{{\tilde P}}
\newcommand{\genjac}{R}
\newcommand{\genjacnmk}{\genjac_{n-k}}
\newcommand{\genjacmmj}{\genjac_{m-j}}
\newcommand{\genjacw}{w_\genjac}
\newcommand{\jacw}{w_\jac}
\newcommand{\normgenjac}{\omega_\genjac}
\newcommand{\normjac}{\omega_\jac}

@@

\title{Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums}
\author{Sheehan Olver, Ben Snowball}
%\date{}							% Activate to display a given date or no date

@@

\begin{abstract}
Sparse spectral methods for solving partial differential equations have been derived in recent years using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices (e.g. a half disk) and trapeziums. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials.  We apply the framework to solving the Poisson, variable coefficient Helmholtz, and Biharmonic equations.
 \end{abstract}

@@

An alternative approach is to use \sotodoinline{Compare with "Discontinuous Galerkin scheme for the spherical shallow water equations with applications to tsunami modeling and prediction"}

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs in cartesian coordinates ($x$, $y$, and $z$) on a half-sphere  can be represented using two families of OPs on the half-disk, see \cite[Theorem 3.1]{olver2018orthogonal} for a similar construction of OPs on an arc in 2D.
%, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. 
Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction \cite{staniforth2012horizontal}. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in turbulent pipe flow.

@@

\noindent \secref{OPs}: We present our procedure  to gain a (two-parameter) family of 2D orthogonal polynomials (OPs) on the half disk domain, by combining 1D OPs on the interval, to form 2D OPs on the disk.

@@

\noindent\secref{Computation}: We discuss computational issues, in particular, how to realise the results of the preceding sections in practice.  We  derive a quadrature rule on the half disk that can be used to expand a function in the OP basis up to a given order.  We implement function evaluation using the coefficients of the expansion of a given function using the Clenshaw algorithm.

@@

\section{Orthogonal polynomials on the half disk}\label{Section:OPs}

@@

For disk slices and trapeziums, we specialise Proposition \ref{prop:construction} in the following definition.
\begin{definition}\label{def:construction}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $\jacw^{(a,b)}(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
\jacw^{(a,b)}(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
and define the associated inner products by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{\jacw^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: \jacw^{(a,b)}(y)\: \D y \label{eqn:ipjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: \jacw^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
\end{align}
Further, denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{\jac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(\jacw^{(a,b)}\) is an even function (i.e. $a = b$, and we can hence denote the weight as $\jacw^{(a)}(x) = w^{(a,a)}_P(x) =  (\delta-x^2)^a$).
\end{enumerate}
We define the four-parameter 2D orthogonal polynomials on $\Omega$ by
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
which are orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: \jacw^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
\end{align*}
that is,
\begin{align*}
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j},
\end{align*}
where for $f, g : \Omega \to \R$ the inner product is defined as 
\begin{align*}
	\ip< f, \: g >_{W^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
\end{align*}
\end{definition}

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align*}
\begin{cases}
(\alpha,\beta) &:= (0,1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= \genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= \jacw^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align*}
Note here we have made the adjustment that $\genjacw^{(a,b,c)}(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of presentiation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

The weight $\jacw(x)$ is an ultraspherical weight, and the corresponding OPs are the normalized Jacobi polynomials $\jac_n^{(b, b)}\}$. The weight $w_R(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation $\{\genjac_n^{(a, b)}\}$ for the orthogonal polynomials. The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{\jac_n^{(b,b)}\}$ are orthonormal, with respect to the inner products respectively given by:
\begin{align*}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_0^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \\
	\ip<p, q>_{\jacw^{(b)}} &:= \frac{1}{\normjac^{(b)}} \: \int_{-1}^1 p(y) \: q(y) \: \jacw^{(b)}(y)\: \D y,
\end{align*}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}} = 1$ $\forall n = 0,1,2,\dots$ and $\norm{\jac_{n}^{(b,b)}}_{\jacw^{(b)}} = 1$ $\forall n = 0,1,2,\dots$. For clarity, we use the notation that $\norm{p}_{\genjacw^{(a,b)}}^2 := \ip<p, p>_{\genjacw^{(a,b)}}$ and that $\norm{p}_{\jacw^{(b)}}^2 := \ip<p, p>_{\jacw^{(b)}}$.

Thus we arrive at a two parameter family of 2D orthogonal polynomials $\{\hdopnkab\}$ given by, for $0 \le k \le n, \: n = 0,1,2,\dots,$
\begin{align}\label{eq:diskpolys}
	\hdopnk^{(a,b)}(x,y) := \genjacnmk^{(a, b+k+\half)}(x) \: \rho(x)^k \: \jac_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on $\Omega$ with respect to the weight
\begin{align*}
	W^{(a,b)}(x,y) = \genjacw^{(a,b)}(x) \: \jacw^{(b)}\fpr(\frac{y}{\rho(x)}) =  \frac{x^a \: (1-x^2-y^2)^b}{\normgenjac^{(a,b)} \normjac^{(b)}}.
\end{align*}
That is, they are orthogonal (for given parameters $a,b$) with respect to the inner product given by
\begin{align*}
	\ip< \hdopnkab, \: \hdopmjab >_\Wab &=  \iint_\Omega \: \hdopnkab(x,y) \: \hdopmjab(x,y) \: \Wab(x,y) \: \D y \: \D x.
\end{align*}
We can see that they are indeed orthogonal using the change of variable $t = \frac{y}{\rho(x)}$:
\begin{align}
	\ip< \hdopnkab, \: \hdopmjab >_\Wab &=  \iint_\Omega \: \Big[ \genjacnmk^{(a,b+k+\half)}(x) \: \genjacmmj^{(a,b+j+\half)}(x) \: \rho(x)^{k+j}  \nonumber \\
		&\quad \quad \quad \quad \quad \quad \cdot \: \jac_k^{(b,b)}\fpr(\frac{y}{\rho(x)}) \: \jac_j^{(b,b)}\fpr(\frac{y}{\rho(x)})  \: \Wab(x,y) \Big] \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x) \: {\genjacmmj^{(a,b+j+\half)}}(x) \: \genjacw^{(a,b+\half(k+j+1))}(x) \: \D x \Big) \nonumber \\
	&\quad \quad \quad \quad \quad \quad \cdot \: \Big( \int_\gamma^\delta \: {\jac_k^{(b,b)}}(t) \:  {\jac_j^{(b,b)}}(t) \: \jacw^{(b)}(t) \: \D t \Big) \nonumber \\
	&= \normjac^{(b)} \: \delta_{k,j} \: \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x) \: {\genjac_{m-k}^{(a,b+k+\half)}}(x) \: \genjacw^{(a,b+k+\half)}(x) \: \D x \nonumber \\ 
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)} \: \delta_{n,m} \: \delta_{k,j}. \label{eqn:normhdop}
\end{align}
Again, for clarity we will use the notation that for $p : \Omega \to \R$, $\norm{p}_{\Wab}^2 := \ip<p, p>_{\Wab}$.

@@

\begin{proof}
The 3-term recurrence for multiplication by $x$ follows from equation (\ref{eqn:Hrecurrence}). For the recurrence for multiplication by $y$, since $\{\hdopmjab\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n+1$ polynomial, we can expand $y \: \hdopnkab(x,) = \sum_{m=0}^{n+1} \sum_{j=0}^m c_{m,j} \: \hdopmjab(x,y)$. These coefficients are given by
\begin{align*}
	c_{m,j} = {\ip< y \: \hdopnkab, \hdopmjab >_{\Wab}}{\norm{\hdopmjab}^{-2}_{\Wab}}.
\end{align*}
Recall from equation (\ref{eqn:normhdop}) that $\norm{\hdopmjab}_{\Wab}^2 = \normgenjac^{(a,b+j+\half)} \: \normjac^{(b)}$. Then for $m = 0,\dots,n+1$, $j = 0,\dots,m$, using the change of variable $t = \frac{y}{\rho(x)}$:
\begin{align*}
	\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \rho(x)^{k+j+2} \: \genjacw^{(a, b)}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} \jac_k^{(b,b)}(t) \: \jac_j^{(b,b)}(t) \: t \: \jacw^{(b)}(t) \: \D t \Big) \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \genjacw^{(a, b+1+\half(k+j))}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} \jac_k^{(b,b)}(t) \: \jac_j^{(b,b)}(t) \: t \: \jacw^{(b)}(t) \: \D t \Big)
\end{align*}
\begin{align*}
	=
	\begin{cases}
    		\delta_k^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k+\frac{3}{2})} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
		\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k-\half)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}} \quad& \text{if } j = k-1 \\
		0 & \text{otherwise}
      	\end{cases}
\end{align*}
where $\ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}}$ and $\ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}}$ are zero for $m < n-1$.

@@

Recall that a function $f(x,y)$ defined on $\Omega$ is approximated by its expansion $f(x,y) = \bighdopab(x,y)^\top \mathbf{f}$. Recall also that $\Wab(x,y) := \genjacw^{(a,b)}(x) \: \jacw^{(b)}(\frac{y}{\rho(x)})$, where $\rho(x) := (1-x^2)^\half$, $\genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b$ and $\jacw^{(b)}(y) := (1-y^2)^b$ for the half disk case.

>>>>

\newcommand{\hdop}{H}
\newcommand{\bighdop}{\mathbb{\hdop}}
\newcommand{\hdopnk}{\hdop_{n,k}}
\newcommand{\hdopab}{\hdop^{(a,b)}}
\newcommand{\hdopnkab}{\hdop_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\hdopmj}{\hdop_{m,j}}
\newcommand{\hdopmjab}{\hdop_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bighdopab}{\bighdop^{(a,b)}}
\newcommand{\Dnt}{D^\top_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\hdopii}{\hdop^{(1,1)}}
\newcommand{\bighdopii}{{\mathbb{\hdop}^{(1,1)}}}
\newcommand{\hdopoo}{\hdop^{(0,0)}}
\newcommand{\bighdopoo}{{\mathbb{\hdop}^{(0,0)}}}
\newcommand{\hdopoi}{\hdop^{(0,1)}}
\newcommand{\hdopio}{\hdop^{(1,0)}}
\newcommand{\jac}{{\tilde P}}
\newcommand{\genjac}{R}
\newcommand{\genjacnmk}{\genjac_{n-k}}
\newcommand{\genjacmmj}{\genjac_{m-j}}
\newcommand{\genjacw}{w_\genjac}
\newcommand{\jacw}{w_P}
\newcommand{\normgenjac}{\omega_\genjac}
\newcommand{\normjac}{\omega_P}

@@

\title{Sparse spectral and $p$-finite element methods for partial differential equations on disk slices and trapeziums}
\author{Sheehan Olver, Ben Snowball}
%\date{}							% Activate to display a given date or no date

@@

\begin{abstract}
Sparse spectral methods for solving partial differential equations have been derived in recent years using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices (e.g. a half-disk) and trapeziums. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials.  We apply the framework to solving the Poisson, variable coefficient Helmholtz, and Biharmonic equations.
 \end{abstract}

@@

%An alternative approach is to use \sotodoinline{Compare with "Discontinuous Galerkin scheme for the spherical shallow water equations with applications to tsunami modeling and prediction"}

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs in cartesian coordinates ($x$, $y$, and $z$) on a half-sphere  can be represented using two families of OPs on the half-disk, see \cite[Theorem 3.1]{olver2018orthogonal} for a similar construction of OPs on an arc in 2D, and it is clear from the construction in this paper that discretizations of spherical gradients and Laplacian's are sparse on half-spheres and other suitable sub-components of the sphere. The resulting sparsity in high-polynomial degree discretisations presents an attractive alternative to methods based on bijective mappings (e.g., \cite{DGShallowWater,FEMShallowWater}).
%
%, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. 
Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction \cite{staniforth2012horizontal}. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in turbulent pipe flow.

@@

\noindent \secref{OPs}: We present our procedure  to gain a (two-parameter) family of 2D orthogonal polynomials (OPs) on the half-disk domain, by combining 1D OPs on the interval, to form 2D OPs on the disk.

@@

\noindent\secref{Computation}: We discuss computational issues, in particular, how to realise the results of the preceding sections in practice.  We  derive a quadrature rule on the half-disk that can be used to expand a function in the OP basis up to a given order.  We implement function evaluation using the coefficients of the expansion of a given function using the Clenshaw algorithm.

@@

\section{Orthogonal polynomials on the half-disk}\label{Section:OPs}

@@

For disk slices and trapeziums, we specialise Proposition \ref{prop:construction} in the following definition. First we introduce notation for two families of univariate OPs
\begin{definition}\label{def:OPconstruction}
Let $\genjacw^{(a,b,c)}(x)$ and $\jacw^{(a,b)}(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
\jacw^{(a,b)}(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
and define the associated inner products by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{\jacw^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: \jacw^{(a,b)}(y)\: \D y \label{eqn:ipjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: \jacw^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
\end{align}
Denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{\jac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).
\end{definition}

\begin{definition}\label{def:constuction}
Define the four-parameter 2D orthogonal polynomials via:
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
\end{definition}

 $\{\hdopnk^{(a,b,c,d)}\}$ are orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: \jacw^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
\end{align*}
assuming that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(\jacw^{(a,b)}\) is an even function (i.e. $a = b$, and we can hence denote the weight as $\jacw^{(a)}(x) = w^{(a,a)}_P(x) =  (\delta-x^2)^a$).
\end{enumerate}
That is,
\begin{align*}
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j},
\end{align*}
where for $f, g : \Omega \to \R$ the inner product is defined as 
\begin{align*}
	\ip< f, \: g >_{W^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
\end{align*}

For the half-disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align*}
\begin{cases}
(\alpha,\beta) &:= (0,1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= \genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= \jacw^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align*}
Note here we make the adjustment that $\genjacw^{(a,b,c)}(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and remove it:
\begin{align}\label{eq:diskpolys}
	\hdopnk^{(a,b)}(x,y) := \genjacnmk^{(a, b+k+\half)}(x) \: \rho(x)^k \: \jac_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega
\end{align}
where $\genjac_n^{(a, b)}  := \genjac_n^{(0,a, b)}$, and we use $W^{(a,b)}(x,y) := W^{(0,a,b,b)}$. In this case the weight $\jacw(x)$ is an ultraspherical weight, and the corresponding OPs are the normalized Jacobi polynomials $\{\jac_n^{(b, b)}\}$, while the weight $w_R(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}).
%
%The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{\jac_n^{(b,b)}\}$ are orthonormal, with respect to the inner products respectively given by:
%\begin{align*}
%	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_0^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \\
%	\ip<p, q>_{\jacw^{(b)}} &:= \frac{1}{\normjac^{(b)}} \: \int_{-1}^1 p(y) \: q(y) \: \jacw^{(b)}(y)\: \D y,
%\end{align*}
%i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}} = 1$ $\forall n = 0,1,2,\dots$ and $\norm{\jac_{n}^{(b,b)}}_{\jacw^{(b)}} = 1$ $\forall n = 0,1,2,\dots$. For clarity, we use the notation that $\norm{p}_{\genjacw^{(a,b)}}^2 := \ip<p, p>_{\genjacw^{(a,b)}}$ and that $\norm{p}_{\jacw^{(b)}}^2 := \ip<p, p>_{\jacw^{(b)}}$.
%
%Thus we arrive at a two parameter family of 2D orthogonal polynomials $\{\hdopnkab\}$ given by, for $0 \le k \le n, \: n = 0,1,2,\dots,$

%orthogonal on $\Omega$ with respect to the weight
%\begin{align*}
%	W^{(a,b)}(x,y) = \genjacw^{(a,b)}(x) \: \jacw^{(b)}\fpr(\frac{y}{\rho(x)}) =  \frac{x^a \: (1-x^2-y^2)^b}{\normgenjac^{(a,b)} \normjac^{(b)}}.
%\end{align*}
%That is, they are orthogonal (for given parameters $a,b$) with respect to the inner product given by
%\begin{align*}
%	\ip< \hdopnkab, \: \hdopmjab >_\Wab &=  \iint_\Omega \: \hdopnkab(x,y) \: \hdopmjab(x,y) \: \Wab(x,y) \: \D y \: \D x.
%\end{align*}
We can see that they are indeed orthogonal using the change of variable $t = \frac{y}{\rho(x)}$, for the following normalisation:
\begin{align}
	\ip< \hdopnkab, \: \hdopmjab >_\Wab &=  \iint_\Omega \: \Big[ \genjacnmk^{(a,b+k+\half)}(x) \: \genjacmmj^{(a,b+j+\half)}(x) \: \rho(x)^{k+j}  \nonumber \\
		&\quad \quad \quad \quad \quad \quad \cdot \: \jac_k^{(b,b)}\fpr(\frac{y}{\rho(x)}) \: \jac_j^{(b,b)}\fpr(\frac{y}{\rho(x)})  \: \Wab(x,y) \Big] \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x) \: {\genjacmmj^{(a,b+j+\half)}}(x) \: \genjacw^{(a,b+\half(k+j+1))}(x) \: \D x \Big) \nonumber \\
	&\quad \quad \quad \quad \quad \quad \cdot \: \Big( \int_\gamma^\delta \: {\jac_k^{(b,b)}}(t) \:  {\jac_j^{(b,b)}}(t) \: \jacw^{(b)}(t) \: \D t \Big) \nonumber \\
	&= \normjac^{(b)} \: \delta_{k,j} \: \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x) \: {\genjac_{m-k}^{(a,b+k+\half)}}(x) \: \genjacw^{(a,b+k+\half)}(x) \: \D x \nonumber \\ 
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)} \: \delta_{n,m} \: \delta_{k,j}. \label{eqn:normhdop}
\end{align}
%Again, for clarity we will use the notation that for $p : \Omega \to \R$, $\norm{p}_{\Wab}^2 := \ip<p, p>_{\Wab}$.

@@

\begin{proof}
The 3-term recurrence for multiplication by $x$ follows from equation (\ref{eqn:Hrecurrence}). For the recurrence for multiplication by $y$, since $\{\hdopmjab\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n+1$ polynomial, we can expand $y \: \hdopnkab(x,y) = \sum_{m=0}^{n+1} \sum_{j=0}^m c_{m,j} \: \hdopmjab(x,y)$. These coefficients are given by
\begin{align*}
	c_{m,j} = {\ip< y \: \hdopnkab, \hdopmjab >_{\Wab}}{\norm{\hdopmjab}^{-2}_{\Wab}}.
\end{align*}
Recall from equation (\ref{eqn:normhdop}) that $\norm{\hdopmjab}_{\Wab}^2 = \normgenjac^{(a,b+j+\half)} \: \normjac^{(b)}$. Then for $m = 0,\dots,n+1$, $j = 0,\dots,m$, using the change of variable $t = \frac{y}{\rho(x)}$:
\begin{align*}
	\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \rho(x)^{k+j+2} \: \genjacw^{(a, b)}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} \jac_k^{(b,b)}(t) \: \jac_j^{(b,b)}(t) \: t \: \jacw^{(b)}(t) \: \D t \Big) \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \genjacw^{(a, b+1+\half(k+j))}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} \jac_k^{(b,b)}(t) \: \jac_j^{(b,b)}(t) \: t \: \jacw^{(b)}(t) \: \D t \Big)
\end{align*}
\begin{align*}
	=
	\begin{cases}
    		\delta_k^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k+\frac{3}{2})} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
		\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k-\half)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}} \quad& \text{if } j = k-1 \\
		0 & \text{otherwise}
      	\end{cases}
\end{align*}
where $\ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}}$ and $\ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}}$ are zero for $m < n-1$.

@@

Recall that a function $f(x,y)$ defined on $\Omega$ is approximated by its expansion $f(x,y) = \bighdopab(x,y)^\top \mathbf{f}$. Recall also that $\Wab(x,y) := \genjacw^{(a,b)}(x) \: \jacw^{(b)}(\frac{y}{\rho(x)})$, where $\rho(x) := (1-x^2)^\half$, $\genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b$ and $\jacw^{(b)}(y) := (1-y^2)^b$ for the half-disk case.
